\textls[-0]{\thetitle}

### 2023-02-28 introduction.tex
tweaks

In the Coq proof assistant~\cite{coq}, for instance, inductive types, as well as functions and
theorems over inductive types, are closed to extension.
So to reuse mechanized metatheories,
the common practice is still to copy code and proofs and then modify them in each extension.
But having to maintain multiple copies is highly non-modular and
antithetical to good software engineering.
%
The programmer could also turn to design patterns and encodings~\cite{delaware2011,delaware2013,schwaab2013modular,keuchel2013generic,forsta2020}.
But they tend to require heavy lifting from the programmer to make code
extensible and often lead to non-idiomatic programming styles.

>>>>

In the \citeauthor{coq} proof assistant, for instance, inductive types, as well as functions and
theorems over inductive types, are closed to extension.
So to reuse mechanized metatheories,
the common practice is still to copy code and proofs and then modify them in each extension.
But having to maintain multiple copies is highly non-modular and
antithetical to good software engineering.
%
The programmer could also turn to design patterns and encodings~\cite{delaware2011,delaware2013,schwaab2013modular,keuchel2013generic,forsta2020}.
But they tend to require heavy lifting from the programmer to make code
extensible and often lead to non-idiomatic programming styles.

### 2023-02-28 metatheory-condensed.tex
tweaks

We contribute \TT, a core type theory that extends
%\underline{f}a\underline{m}i\underline{l}y polymorphism
%\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory (MLTT)~\cite{martin1982constructive}
\citeauthor{martin1982constructive} dependent type theory (MLTT)
with facilities to express family polymorphism
while maintaining consistency and canonicity.
%\NA{maybe you can give an overview of the salient points of the design and what they're trying to achieve? What is the core aspect modeled by the theory?}
%
We acknowledge that the presentation herein is dense for an audience
without prior exposure to MLTT, so we summarize the salient points
first before delving into the technical details.

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT
(and \cref{fig:fmltt-selected} \TT).
%We briefly review the base MLTT fragment first.
%
Following the current trend~\cite{coquand2018canonicity,altkap2016,gratzer-sterling-birkedal-2019}
in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

>>>>

We contribute \TT, a core type theory that extends
%\underline{f}a\underline{m}i\underline{l}y polymorphism
%\underline{M}artin-\underline{L}öf dependent \underline{t}ype \underline{t}heory (MLTT)~\cite{martin1982constructive}
\citeauthor{martin1982constructive} dependent type theory (MLTT)
with facilities to express family polymorphism
while maintaining consistency and canonicity.
%\NA{maybe you can give an overview of the salient points of the design and what they're trying to achieve? What is the core aspect modeled by the theory?}
%
We acknowledge that the presentation herein is dense for an audience
unacquainted with MLTT or explicit substitutions, so we summarize the salient points
first before delving into the technical details.

@@

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT
(and \cref{fig:fmltt-selected} \TT).
%We briefly review the base MLTT fragment first.
%
Following the current trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}
in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
of the syntax rather than meta-operations.
%
Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).
